Predicting prolonged apneic episodes using machine learning during nurse-administered procedural sedation

### 2021-03-12 manuscript.Rmd
combined plot

A potential consequence to consider in regard to the implementation of capnography for patient monitoring in inpatient wards is alarm fatigue. Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successful implementation of this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation. presumably in an attempt to reduce alarm burden, capnography monitor alarms now mostly default to triggering after apnea has been observed for 30 seconds. It would be ideal for clinicians to receive capnography monitor alarms as *_early as possible_* in the course of an apneic event, *but only* if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect.

@@

Based on results from prior research, it seems that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods. It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@Conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states. Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department. It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).

We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach. The aim was to determine the accuracy of machine learning models for predicting if a period of apnea will persist for 30 seconds or more.

@@

Net benefit associated with the random forest model exceeded the strategy of intervening for all or no patients across all probability thresholds from the range of 0.4 to 0.8. The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was 40%, would be achieved by using the random forest model as opposed to treating everyone or no one.o

>>>>

A potential consequence to consider in regard to the implementation of capnography for patient monitoring in inpatient wards is alarm fatigue. Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successful implementation of this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation. Presumably in an attempt to reduce alarm burden, capnography monitor alarms now mostly default to triggering after apnea has been observed for 30 seconds. It would be ideal for clinicians to receive capnography monitor alarms as *_early as possible_* in the course of an apneic event, *but only* if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect.

@@

The aim was to determine the accuracy of machine learning models for predicting if a period of apnea will persist for 30 seconds or more.

@@

Net benefit associated with the random forest model exceeded the strategy of intervening for all patients across all probability thresholds from the range of 0.4 to 0.8. The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40%, would be achieved by using the random forest model as opposed to treating everyone. In contrast, the net benefit associated with the random forest model was lower than the strategy of not intervening for any patients across all probability thresholds from the range of 0.4 to 0.8. This was due to the fact that we applied a condition that we would not accept more than 5 false positives to detect 1 true positive. The number of false positives that triggered across probability thresholds from 0.4 to 0.8 using the random forest model would far exceed the number of true positives, meaning the preferred strategy should remain as only setting the alert for the 30-second time-point.

## Discussion

Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods. It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@Conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states. Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department. It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).

We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach.

Using the machine learning model would produce a better clinical outcome, in terms of the balance between true positives and false positives for detection of apneic periods over 30 seconds, than a default strategy of intervening for everyone who has an apneic period ≥15 seconds.
